Process-based phenotypes: an application to the Cellular Phenotype Ontology

### 2011-11-22 paper/drinking.tex

Single occurrences of processes can be abnormal in multiple ways,
depending on the type of process.
%
First, common to all processes is the quality of {\em duration} and
consequently, each process can have an {\em abnormal} (increased or
decreased) duration. The second common type of abnormality are
abnormalities based on process participants in relation to the
duration of the process. These include all kinds of {\em rates} such
as {\em mass flow rate}, {\em energy flow rate} and {\em velocity}
(the rate of change of position). In each of these cases, an object
participates in a process and a quality (or change of quality) of that
object throughout the duration of the process is considered to form a
new quality. If the process has participants that are distinguished
into {\em inputs} and {\em outputs}, then a recurring pattern is that
the amount of inputs or outputs that participate in the process can be
{\em increased} or {\em decreased}.  Finally, some objects may be
divided into stages during which particular states of affairs obtain,
and a process may be abnormal in that these states of affairs do not
obtain at a particular stage. Notably, at the beginning and the end of
a process, pre- and post-conditions may obtain that are abnormally
changed in a process. For example, {\em Aneuploidy} -- an abnormality
during cell division at which the chromosomes do not separate properly
between the two cells -- may be considered the result of such an
abnormality.

@@

A second kind of regulatory abnormalities is related to the {\em
  onset} of a process. With respect to a reference process, a
particular kind of process may be {\em delayed} ({\tt PATO:0000502})
or {\em premature} ({\tt PATO:0000694}). For example, {\em Delayed
  apoptosis} refers to an abnormality of the {\em Regulation of
  apoptosis} in which apoptosis is induced later than normal.
We use

>>>>

Single occurrences of processes can be abnormal in multiple ways,
depending on the type of process.
%
First, common to all processes is the quality of {\em duration} and
consequently, each process can have an {\em abnormal} (increased or
decreased) duration. For example, a part of an organism may
participate in an {\em Inflammatory response} ({\tt GO:0006954}) that
lasts longer than normal, i.e., the organism has an {\em Increased
  duration of inflammatory response} phenotype. We define such a
phenotype as a phenotype of an organism which has a part that
participates in {\em Inflammatory response}, and this {\em
  Inflammatory response} process has an {\em Increased duration} ({\tt
  PATO:0000498}).

The second common type of abnormality are abnormalities based on
process participants in relation to the duration of the process. These
include all kinds of {\em rates} such as {\em mass flow rate}, {\em
  energy flow rate} and {\em velocity} (the rate of change of
position). In each of these cases, an object participates in a process
and a quality (or change of quality) of that object throughout the
duration of the process is considered to form a new quality. If the
process has participants that are distinguished into {\em inputs} and
{\em outputs}, then a recurring pattern is that the amount of inputs
or outputs that participate in the process can be {\em increased} or
{\em decreased}. For example, an {\em Increased rate of cytoplasmic
  streaming} is defined as

Finally, some objects may be divided into stages during which
particular states of affairs obtain, and a process may be abnormal in
that these states of affairs do not obtain at a particular
stage. Notably, at the beginning and the end of a process, pre- and
post-conditions may obtain that are abnormally changed in a
process. For example, {\em Aneuploidy} -- an abnormality during cell
division at which the chromosomes do not separate properly between the
two cells -- may be considered the result of such an abnormality.

@@

A second kind of regulatory abnormalities is related to the {\em
  onset} of a process. With respect to a reference process, a
particular kind of process may be {\em delayed} ({\tt PATO:0000502})
or {\em premature} ({\tt PATO:0000694}). For example, {\em Delayed
  apoptosis} refers to an abnormality of the {\em Regulation of
  apoptosis} in which apoptosis is induced later than normal.  We use
the PATO quality {\em onset} ({\tt PATO:...}) and its children {\em
  delayed} and {\em premature} to define these types of regulatory
abnormality.

Finally, a third kind of regulatory abnormality refers to abnormal
rates with respect to a participant of the process that is being
regulated.
